---
title: "ANOVA, glm and LME analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(vegan)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(car)
library(margins)
library(gridExtra)
library(broom)
library(rsq) 
library("lmerTest")
```

# load and manipulate data
```{r}
metadata_file <- read_csv("data/metadata_timeseries_dereplicated.csv", 
                          col_types = cols_only(
                            Study = col_factor(), 
                            Environment = col_factor(), 
                            Gut_sample_type = col_factor(),
                            Host = col_factor(),
                            Age_group = col_factor(), 
                            Age_months = col_character(), 
                            Status = col_factor(), 
                            Premature = col_factor(),
                            Delivery_mode = col_factor(),
                            Geography = col_factor(), 
                            Sex = col_factor(),
                            Abx = col_factor(), 
                            CDI = col_factor(), 
                            Cdiff = col_factor(),
                            final_representative = col_character(), 
                            bmi = col_double(), 
                            diet = col_factor(), 
                            total_MG_coverage = col_double()
))

#preliminary filtering
metadata_file2 <- metadata_file %>%
  filter(Host=="Human" & Environment=="Gut" & Gut_sample_type=="Stool" & final_representative=="Yes") %>% 
  mutate(Cdiff_binary = ifelse(Cdiff == "Yes", 1, 0),
         bmi_categories = case_when(
             (bmi<18.5) ~ "underweight",
             (bmi>=18.5 & bmi <= 24.9) ~ "normal",
             (bmi>=25 & bmi <= 29.9) ~ "overweight",
             TRUE ~ "obese"), 
         diet_infant = ifelse((diet != "omnivorous" & diet != "vegan" & diet != "vegetarian"), diet, NA))

metadata_file2$bmi_categories <- as.factor(metadata_file2$bmi_categories) 
metadata_file2$diet_infant <- as.factor(metadata_file2$diet_infant) 

```

# sequential ANOVA divided per age group (Infants and adults only)
```{r}

#calculate sequential ANOVA for infants
metadata_file2.infant <- metadata_file2 %>% 
  filter(Age_group=="Infant")

aov.cumulative.infant <- 
  Anova(glm(Cdiff_binary ~ total_MG_coverage + Geography + Abx + Sex + Status + Premature + Delivery_mode + diet, family = "binomial", data=metadata_file2.infant), type=2) %>%
  broom::tidy()

aov.cumulative.infant$p.adj <- p.adjust(aov.cumulative.infant$p.value)
#print(aov.cumulative.infant)

#calculate sequential ANOVA for adults
metadata_file2.adult <- metadata_file2 %>% 
  filter(Age_group=="Adult")

aov.cumulative.adult <- 
  Anova(glm(Cdiff_binary ~ total_MG_coverage + Geography + Abx + Sex + Status + bmi_categories, 
            family = "binomial", data=metadata_file2.adult), type=2) %>%
  broom::tidy()

aov.cumulative.adult$p.adj <- p.adjust(aov.cumulative.adult$p.value)
#print(aov.cumulative.adult)

aov.cumulative.infant.intervals <- aov.cumulative.infant %>% 
  mutate(significance.padj = case_when(
             (p.adj > 0.05) ~ "ns",
             (p.adj <= 0.05 & p.adj > 0.01) ~ "*",
             (p.adj <= 0.01 & p.adj > 0.001) ~ "**",
             (p.adj <= 0.001) ~ "***",
             TRUE ~ "value out of ranges"),
         type = "sequential_infant", 
         metric = "Cdiff_binary", 
         site = "stool") 


aov.cumulative.adult.intervals <- aov.cumulative.adult %>% 
  mutate(significance.padj = ifelse(p.adj > 0.05, "ns", 
                               ifelse(p.adj <= 0.05 & p.adj > 0.01, "*",
                                      ifelse(p.adj <= 0.01 & p.adj > 0.001, "**",
                                             ifelse(p.adj <= 0.001, "***", "value out of ranges")))),
         type = "sequential_adult", 
         metric = "Cdiff_binary", 
         site = "stool") 
```

#single factors divided by age group (infants and adults only) - combined with Marker Gene (MG) coverage
```{r}
#preliminary check to see which variables should be dropped (less than 2 factors left after removing NAs)
ifelse(n <- sapply(metadata_file2, function(x) length(levels(x))) ==1, "Drop this variable", "OK")

list_continuous1 <-  c("total_MG_coverage")
list_continuous2 <-  c("bmi")
idx_infant = c("Abx", "Sex", "Status", "Premature", "Delivery_mode", "Geography", "diet", "total_MG_coverage")
idx_adult = c("Abx", "Sex", "Status", "bmi_categories", "Geography", "CDI", "total_MG_coverage")

collect.single.infant <- data.frame()
collect.single.adult <- data.frame()

glm.single.rsq.adj <- data.frame() #collected adjusted Rsquared

#calculate single factor glm for infants
for (i in idx_infant) {
  metadata_file2.infants <- metadata_file2 %>%
    filter(Age_group=="Infant")
  
  if (i %in% list_continuous1) { #continuous variables (MG coverage)
    glm.single <- glm(substitute(Cdiff_binary~ i, list(i = as.name(i))), 
                      family = "binomial", 
                      data = metadata_file2.infants, 
                      na.action=na.omit)
  }else{
    if (i %in% list_continuous2) { #continuous variables (BMI)
      glm.single <- glm(substitute(Cdiff_binary~ total_MG_coverage+i, list(i = as.name(i))), 
                      family = "binomial", 
                      data = metadata_file2.infants, 
                      na.action=na.omit)
      } else { # discrete variables
        glm.single <- glm(substitute(Cdiff_binary~ total_MG_coverage + as.factor(i), list(i = as.name(i))), family = "binomial", 
                      data = metadata_file2.infants, na.action=na.omit)
      }
  }
  
  aov.single <- Anova(glm.single) %>% broom::tidy() %>% 
    mutate(site = "human_stool", metric = "Cdiff_binary")
  aov.single <- aov.single[1,]
  aov.single$term <- i
  # collect data
  collect.single.infant <- rbind(collect.single.infant, aov.single)
  
  #coefficients and adjusted R squared
  summary(glm.single)
  adj.rsq <- rsq(glm.single, adj = TRUE) 
  glm.single.rsq.adj <- rbind(glm.single.rsq.adj, data.frame(term = paste0("total_MG_coverage+",i), age = "Infant", rsquared_adjusted = adj.rsq)) # get adjusted R squared
}

#calculate single factor glm for adults
for (i in idx_adult) {
  metadata_file2.adults <- metadata_file2 %>%
    filter(Age_group=="Adult")
  
  if (i %in% list_continuous1) { #continuous variables (MG coverage)
    glm.single <- glm(substitute(Cdiff_binary~ i, list(i = as.name(i))), 
                      family = "binomial", 
                      data = metadata_file2.adults, 
                      na.action=na.omit)
  }else{
    if (i %in% list_continuous2) { #continuous variables (BMI)
      glm.single <- glm(substitute(Cdiff_binary~ total_MG_coverage + i, list(i = as.name(i))), 
                      family = "binomial", 
                      data = metadata_file2.adults, 
                      na.action=na.omit)
      } else { # discrete variables
        glm.single <- glm(substitute(Cdiff_binary~ total_MG_coverage + as.factor(i), list(i = as.name(i))), family = "binomial", 
                      data = metadata_file2.adults, na.action=na.omit)
      }
  }
  
  aov.single <- Anova(glm.single) %>% broom::tidy() %>% 
    mutate(site = "human_stool", metric = "Cdiff_binary")
  aov.single <- aov.single[1,]
  aov.single$term <- i
  # collect data
  collect.single.adult <- rbind(collect.single.adult, aov.single)
  
  #coefficients and adjusted R squared
  summary(glm.single)
  adj.rsq <- rsq(glm.single, adj = TRUE) 
  glm.single.rsq.adj <- rbind(glm.single.rsq.adj, data.frame(term = paste0("total_MG_coverage+",i), age = "Adult", rsquared_adjusted = adj.rsq)) # get adjusted R squared
}

#fdr correction 
collect.single.infant$p.adj <- p.adjust(collect.single.infant$p.value)
collect.single.adult$p.adj <- p.adjust(collect.single.adult$p.value)

glm.single.rsq.adj

#add significant symbols to infant table
collect.single.infant.intervals <- collect.single.infant %>% 
  mutate(significance.padj = case_when(
    (p.adj > 0.05) ~ "ns", 
    (p.adj <= 0.05 & p.adj > 0.01) ~ "*",
    (p.adj <= 0.01 & p.adj > 0.001) ~ "**",
    (p.adj <= 0.001)  ~ "***",
    TRUE ~ "value out of ranges"),
    type = "single_factors_infant")

#add significant symbols to adult table
collect.single.adult.intervals <- collect.single.adult %>% 
  mutate(significance.padj = case_when(
    (p.adj > 0.05) ~ "ns", 
    (p.adj <= 0.05 & p.adj > 0.01) ~ "*",
    (p.adj <= 0.01 & p.adj > 0.001) ~ "**",
    (p.adj <= 0.001)  ~ "***",
    TRUE ~ "value out of ranges"),
    type = "single_factors_adult")


```

#combine single factors and sequential for infants and adults
```{r}

combined.allAges <- rbind(collect.single.infant.intervals,
                          collect.single.adult.intervals,
                          aov.cumulative.infant.intervals,
                          aov.cumulative.adult.intervals)

combined.allAges <-  combined.allAges %>% 
  mutate(term=fct_relevel(term, "total_MG_coverage", "Geography", "Abx", "Sex", "Status", "Premature", "Delivery_mode", "bmi_categories", "diet", "CDI"))

heatmap.combined.allAges.padj <- ggplot(combined.allAges %>% filter(!is.na(p.adj)), aes(term, type, fill= significance.padj)) + 
  geom_tile() +
  theme_bw() +
  scale_fill_manual(values=c("#B2DFDB", "#26A69A", "#004D40", "#EEEEEE", "#EEEEEE")) + #for total_MG_coverage
  theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust = 1.0, size = 10))

```
#sequential glm ANOVA
```{r}
#CDI + bmi + diet + smoker + Birth_weight
glm.cumulative <- glm(Cdiff_binary ~ total_MG_coverage + Geography + Abx + Sex + Age_group + Status + Premature + Delivery_mode, family = "binomial", data=metadata_file2)
aov.cumulative <- Anova(glm.cumulative, type=2) %>% broom::tidy()

aov.cumulative$p.adj <- p.adjust(aov.cumulative$p.value)
print(aov.cumulative)

#get coefficients
summary(glm.cumulative)

#rsq(glm.cumulative) # get R squared
glm.cumulative.rsq.adj <- rsq(glm.cumulative, adj = TRUE) # get adjusted R squared

#calculate sequential glm for infants  
glm.cumulative.infant <- glm(Cdiff_binary ~ total_MG_coverage + Geography + Abx + Sex + Status + Premature + Delivery_mode, 
                             family = "binomial", data=metadata_file2 %>% filter(Age_group=="Infant"))
aov.cumulative.infant <- Anova(glm.cumulative.infant, type=2) %>% broom::tidy()
glm.cumulative.r.infant.sq.adj <- rsq(glm.cumulative.infant, adj = TRUE) # get adjusted R squared

#calculate sequential glm for adults 
glm.cumulative.adult <- glm(Cdiff_binary ~ total_MG_coverage + Geography + Abx + Sex + Status, 
                            family = "binomial", data=metadata_file2 %>% filter(Age_group=="Adult"))
aov.cumulative.adult <- Anova(glm.cumulative.adult, type=2) %>% broom::tidy()
glm.cumulative.r.adult.sq.adj <- rsq(glm.cumulative.adult, adj = TRUE) # get adjusted R squared
```

#save plots
```{r}
pdf(file=paste0("figures/ANOVA/heatmap_padj.pdf"), width=10, height=8)
heatmap.combined.allAges.padj
dev.off()
```

#single factors divided by age group (infants and adults only) - combined with MG coverage - with ODDS RATIO
```{r}
#preliminary check to see which variables should be dropped (less than 2 factors left after removing NAs)
ifelse(n <- sapply(metadata_file2, function(x) length(levels(x))) ==1, "Drop this variable", "OK")

list_continuous1 <-  c("total_MG_coverage")
list_continuous2 <-  c("bmi")
idx_infant = c("Abx", "Sex", "Status", "Premature", "Delivery_mode", "Geography", "diet", "total_MG_coverage")
idx_adult = c("Abx", "Sex", "Status", "bmi_categories", "Geography", "CDI", "total_MG_coverage")

collect.single.infant.or <- data.frame()
collect.single.infant.or <- data.frame()

#infants
for (i in idx_infant) {
  metadata_file2.infants <- metadata_file2 %>%
    filter(Age_group=="Infant")
  
  if (i %in% list_continuous1) { #continuous variables (MG coverage)
    glm.single <- glm(substitute(Cdiff_binary~ i, list(i = as.name(i))), 
                      family = "binomial", 
                      data = metadata_file2.infants, 
                      na.action=na.omit)
  }else{
    if (i %in% list_continuous2) { #continuous variables (BMI)
      glm.single <- glm(substitute(Cdiff_binary~ total_MG_coverage+i, list(i = as.name(i))), 
                      family = "binomial", 
                      data = metadata_file2.infants, 
                      na.action=na.omit)
      } else { # discrete variables
        glm.single <- glm(substitute(Cdiff_binary~ total_MG_coverage + as.factor(i), list(i = as.name(i))), family = "binomial", 
                      data = metadata_file2.infants, na.action=na.omit)
      }
  }
  
  #odds ratio from coefficients
  oddsRatio.single.infants <- glm.single$coefficients[2] %>% broom::tidy()
  oddsRatio.single.infants <- oddsRatio.single.infants[1,]
  oddsRatio.single.infants$term <- i
  # collect data
  collect.single.infant.or <- rbind(collect.single.infant.or, oddsRatio.single.infants)
}

collect.single.adult.or <- data.frame()
collect.single.adult.or <- data.frame()

#adults
for (i in idx_adult) {
  metadata_file2.adults <- metadata_file2 %>%
    filter(Age_group=="Adult")

  if (i %in% list_continuous1) { #continuous variables (MG coverage)
    glm.single <- glm(substitute(Cdiff_binary~ i, list(i = as.name(i))),
                      family = "binomial",
                      data = metadata_file2.adults,
                      na.action=na.omit)
  }else{
    if (i %in% list_continuous2) { #continuous variables (BMI)
      glm.single <- glm(substitute(Cdiff_binary~ total_MG_coverage + i, list(i = as.name(i))),
                      family = "binomial",
                      data = metadata_file2.adults,
                      na.action=na.omit)
      } else { # discrete variables
        glm.single <- glm(substitute(Cdiff_binary~ total_MG_coverage + as.factor(i), list(i = as.name(i))), family = "binomial",
                      data = metadata_file2.adults, na.action=na.omit)
      }
  }

  #odds ratio from coefficients
  oddsRatio.single.adult <- glm.single$coefficients[2] %>% broom::tidy()
  oddsRatio.single.adult <- oddsRatio.single.adult[1,]
  oddsRatio.single.adult$term <- i
  # collect data
  collect.single.adult.or <- rbind(collect.single.adult.or, oddsRatio.single.adult)
}
```

#LME to test association between Cdiff and total MG coverage
```{r}
#CDI data
metadata_file_lme_cdi <- read_csv("data/CDI/CDI_and_splittedControl_samples.csv")
fit_cdi <- lmer(total_MG_coverage~Cdiff+(1|Study), data=metadata_file_lme_cdi)
summary(fit_cdi)

#global data
metadata_file <- metadata_file %>% 
  filter(final_representative=="Yes")

fit <- lmer(total_MG_coverage~Cdiff+(1|Study), data=metadata_file)
summary(fit)
```

#LME cdiff proportion (suppl. figure 4B)
```{r}
library("tidyverse")
library("lmerTest")

temp <- read_csv('data/metadata_timeseries_dereplicated.csv') %>% 
  filter(final_representative=='Yes') %>% 
  filter(!is.na(total_MG_coverage)) %>% 
  mutate(quant=ntile(total_MG_coverage, n = 10)) 

fit <- lmer(as.numeric(as.factor(Cdiff))~quant+(1|Study), data=temp)
summary(fit)
# Fixed effects:
# Estimate Std. Error         df t value Pr(>|t|)    
# (Intercept)  1.072e+00  1.057e-02  3.465e+02 101.464   <2e-16 ***
# quant       -3.682e-04  7.639e-04  2.302e+04  -0.482     0.63 

g <- temp %>% 
  group_by(quant, Cdiff) %>% 
  tally() %>% 
  mutate(n.all=sum(n)) %>% 
  mutate(prop=n/n.all) %>% 
  ggplot(aes(x=as.factor(quant), y=prop, fill=Cdiff)) + 
  geom_bar(stat='identity') + 
  scale_fill_manual(values=c('#A8A99E', '#D41645'), guide='none') + 
  xlab('Coverage quantile') +
  ylab('Proportion') +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid = element_blank()) + 
  NULL
ggsave(g, filename = 'figures/ANOVA/Cdiff_proportion_all.pdf', width = 5, height = 3)

```

