---
title: "first appearance of cdifficile in infants timeseries analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
library("grid")
library(gridExtra)
library(rstatix)
```

#Read metadata Table 
```{r}
metadata_file <- read_csv("data/metadata_timeseries_dereplicated.csv", 
                          col_types = cols_only(
                            Samples = col_character(),
                            Study = col_factor(), 
                            Environment = col_factor(), 
                            Gut_sample_type = col_factor(),
                            Host = col_factor(),
                            Age_group = col_factor(), 
                            Age_months = col_character(), 
                            Status = col_factor(), 
                            Premature = col_factor(),
                            Delivery_mode = col_factor(),
                            CDI = col_factor(), 
                            Cdiff = col_factor(),
                            subject_id = col_character(),
                            unique_subject_id = col_character(),
                            timepoint = col_double(),
                            final_representative = col_character(), 
                            diet = col_character()
))

  
#to order properly by age, this field needs to be numeric
metadata_file$Age_months <- as.double(metadata_file$Age_months)

#identify the first time cdiff appears in longitudinal data per each subject
#notice that here we do NOT filter by final_representative=="Yes", meaning we use all available samples in each subject's timeseries
metadata_file2 <- metadata_file %>%
  filter(Host=="Human" & Environment=="Gut" & Gut_sample_type=="Stool" & !is.na(Age_months) & Age_months <= 84 & !is.na(unique_subject_id)) %>% #limit plot to max 7 years of age
  group_by(unique_subject_id) %>% 
  mutate(n_cdiffPos = sum(Cdiff == "Yes"), 
           n_cdiffNeg = sum(Cdiff == "No"), 
           n_samples = n(),
           #define cases
           #mixed timeseries with cdiff+ and cdiff- samples
           timeseries_category = case_when(
             (n_cdiffPos>0 & n_cdiffNeg>0 & n_samples>1) ~ "mixed", 
             #timeseries with all samples cdiff+
             (n_cdiffPos>0 & n_cdiffNeg==0 & n_samples>1) ~ "all_cdiff_positive", 
             #timeseries with all samples cdiff-
             (n_cdiffPos==0 & n_cdiffNeg>0 & n_samples>1) ~ "all_cdiff_negative", 
             # only one sample per subject
             TRUE ~ "not_timeseries")) %>% 
  #keep only cdiff+ samples from timeseries that had at least 1 cdiff+ sample
  filter(timeseries_category != "all_cdiff_negative" & timeseries_category != "not_timeseries"& Cdiff == "Yes") %>%
  ungroup() %>%
  group_by(unique_subject_id) %>% 
  #sort timepoints within each subject
  dplyr::arrange(Age_months, .by_group = TRUE) %>% 
  #take the first line/appearance
  slice(1)
```

#define age ranges of interest
```{r}
infant_interval = 2 #months
child_adol_interval = 12*3 #3yrs
adult_elderly_interval = 12*10 #10yrs

age_range_as_label <- function(age_range) {
  return(paste0(age_range[1], "_", age_range[2], "M"))
}

replace_age_range <- function(age_ranges, i, sequence, interval) {
  age_ranges[[length(age_ranges)+ 1]] <- c(i, i+interval)
  return(age_ranges)
}

age_ranges <- list()
#ad hoc age ranges 0-1M and 1-2M
age_ranges[[length(age_ranges)+ 1]] <- c(0, 1)
age_ranges[[length(age_ranges)+ 1]] <- c(1, 2)

#infants age ranges defined by equal time intervals
seq_infant <- seq(2, 12-infant_interval, by=infant_interval)
for (i in seq_infant){
  age_ranges <- replace_age_range(age_ranges, i, seq_infant, infant_interval)
}

#children and adolescents
#limit is 228 instead of 216 because 228 is the closest multiple available
seq_child_adol <- seq(12, 228-child_adol_interval, by=child_adol_interval)
for (i in seq_child_adol){
  age_ranges <- replace_age_range(age_ranges, i, seq_child_adol, child_adol_interval)
}

#adults and elderly
seq_adult_elderly <- seq(228, 1308-adult_elderly_interval, by=adult_elderly_interval)
for (i in seq_adult_elderly){
  age_ranges <- replace_age_range(age_ranges, i, seq_adult_elderly, adult_elderly_interval)
}

age_ranges_labels <- mapply(age_range_as_label, age_ranges)
age_ranges_labels
```

#manipulate metadata table
```{r}
convert_range <- function(val, age_ranges) {
  values <- str_split(val, "-")
  # If we have an age range
  if(length(values) == 2) {
    values <- as.double(values)
    for (age_range in age_ranges){
      if (values[1] >= age_range[1] && values[1] < age_range[2] && 
          values[2] >= age_range[1] && values[2] < age_range[2]) {
        return(age_range_as_label(age_range))
      }
    }
  }
  
  # if the age range didn't fit into any category it returns the age interval itself
  if(str_detect(val, "-")) {return(val)}
  val <- as.double(val)
  # if(is.na(val)) {return(val)}
  for (age_range in age_ranges){
    if (val >= age_range[1] && val < age_range[2]) {
      return(age_range_as_label(age_range))
    }
  }
  return(val)
}

convert_to_ranges <- function(age_months, age_ranges){
  mapply(convert_range, age_months, MoreArgs=list(age_ranges=age_ranges))
}

metadata_file_age_ranges <- metadata_file2 %>% 
  mutate(age_plot_range = convert_to_ranges(Age_months, age_ranges))
```

#calculate number samples available per age range
```{r}
metadata_file_total_forPercentage <- metadata_file %>%
  filter(Host=="Human" & Environment=="Gut" & Gut_sample_type=="Stool" & Age_months != "NA" & Age_months <= 84 & unique_subject_id != "NA") %>% #limit plot to max 7 years of age
  group_by(unique_subject_id) %>% 
  mutate(n_cdiffPos = sum(Cdiff == "Yes"), 
           n_cdiffNeg = sum(Cdiff == "No"), 
           n_samples = n(),
           #define cases
           #mixed timeseries with cdiff+ and cdiff- samples
           timeseries_category = case_when(
             (n_cdiffPos>0 & n_cdiffNeg>0 & n_samples>1) ~ "mixed", 
             #timeseries with all samples cdiff+
             (n_cdiffPos>0 & n_cdiffNeg==0 & n_samples>1) ~ "all_cdiff_positive", 
             #timeseries with all samples cdiff-
             (n_cdiffPos==0 & n_cdiffNeg>0 & n_samples>1) ~ "all_cdiff_negative", 
             # only one sample per subject
             TRUE ~ "not_timeseries")) %>% 
  filter(timeseries_category != "not_timeseries") %>% 
  ungroup()

metadata_file_total_forPercentage_ageRanges <- metadata_file_total_forPercentage %>% 
  mutate(age_plot_range = convert_to_ranges(Age_months, age_ranges)) %>% 
  group_by(age_plot_range) %>% 
  dplyr::summarise(total_forPercentage = dplyr::n()) 
```

#color palette for all lines in one plot
```{r}
myPalette <- c("#b35806","#f1a340","#fee0b6","#d8daeb","#998ec3","#542788","green", "red","blue", "#333333")
```


#all lines in one plot
```{r}
metadata_file_age_ranges.all <- metadata_file_age_ranges %>% 
  group_by(age_plot_range) %>% 
  dplyr::summarise(occurrences = dplyr::n()) %>% 
  mutate(Other_variable = "total")

metadata_file_age_ranges.Premature <- metadata_file_age_ranges %>% 
  filter(Premature != "NA") %>% 
  group_by(age_plot_range, Premature) %>% 
  dplyr::summarise(occurrences = dplyr::n()) %>% 
  mutate(Other_variable = ifelse(Premature == "Yes", "Pre-term", "Full-term"))

metadata_file_age_ranges.Status <- metadata_file_age_ranges %>% 
  filter(Status != "NA") %>% 
  group_by(age_plot_range, Status) %>% 
  dplyr::summarise(occurrences = dplyr::n()) %>% 
  mutate(Other_variable = Status)

metadata_file_age_ranges.DeliveryMode <- metadata_file_age_ranges %>% 
  filter(Delivery_mode != "NA") %>% 
  group_by(age_plot_range, Delivery_mode) %>% 
  dplyr::summarise(occurrences = dplyr::n()) %>% 
  mutate(Other_variable = ifelse(Delivery_mode == "Vaginal", "Vaginal delivery", "C-section delivery"))

metadata_file_age_ranges.Diet <- metadata_file_age_ranges %>% 
  filter(diet != "NA") %>% 
  group_by(age_plot_range, diet) %>% 
  dplyr::summarise(occurrences = dplyr::n()) %>% 
  mutate(Other_variable = ifelse(diet == "mixed feeding", "mixed_bottlefed_breastfed", diet))
```

#plot totals line with status line
```{r}
metadata_file_age_ranges.allLines.status <- bind_rows(metadata_file_age_ranges.all, metadata_file_age_ranges.Status)

# total_forPercentage provides the total number of samples available oer age range without specific filters (i.e. both cdiff+ and cdiff- timeseries are included)
metadata_file_age_ranges.allLines_total_forPercentage.status <- left_join(metadata_file_age_ranges.allLines.status,  metadata_file_total_forPercentage_ageRanges)
#divide by total_forPercentage to get prevalence in %
metadata_file_age_ranges.allLines_total_forPercentage.status <- metadata_file_age_ranges.allLines_total_forPercentage.status %>% 
  mutate(percentage = occurrences*100/total_forPercentage)


metadata_file_age_ranges.allLines_total_forPercentage.status <- metadata_file_age_ranges.allLines_total_forPercentage.status %>% 
  mutate(Other_variable = fct_relevel(Other_variable, "Total", "Healthy", "Diseased"),
         point_size = case_when(
             (total_forPercentage>=5 & total_forPercentage<50) ~ 1, 
             (total_forPercentage>=50 & total_forPercentage<100) ~ 2,
             (total_forPercentage>=100 & total_forPercentage<500) ~ 3,
             (total_forPercentage>=500) ~ 4,
             #less than n=5 observations supporting the group
             TRUE ~ 0))

readable_age_label <- c("0-1 mos","1-2 mos","2-4 mos","4-6 mos","6-8 mos","8-10 mos","10-12 mos","1-4 yrs","4-7 yrs")

p.status<- ggplot(data=metadata_file_age_ranges.allLines_total_forPercentage.status, aes(x=factor(age_plot_range, levels = c("0_1M","1_2M","2_4M","4_6M","6_8M","8_10M","10_12M","12_48M","48_84M","84_120M")), y=percentage, group=Other_variable)) +
  geom_line(aes(color=Other_variable, linetype=Other_variable), size = 1.2) +
  scale_linetype_manual(values=c("solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid"))+
  scale_color_manual(values=myPalette) +
  geom_point(pch = 21, color="black", stroke = 1, aes(fill=Other_variable, size = as.factor(point_size))) + 
  scale_fill_manual(values=myPalette) +
  scale_size_discrete(labels=c("100 \u2264 N. samples < 500", "N. samples \u2265  500")) +
  labs(x="Age groups", y = "Prevalence %", size = 10)+
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust = 1.0)) +
  scale_x_discrete(labels= readable_age_label) +
  theme(legend.title=element_blank(),
  legend.position = "none")

p.status
```

#plot totals line with delivery mode line
```{r}
metadata_file_age_ranges.allLines.delivery <- bind_rows(metadata_file_age_ranges.all, metadata_file_age_ranges.DeliveryMode)

# total_forPercentage provides the total number of samples available oer age range without specific filters (i.e. both cdiff+ and cdiff- timeseries are included)
metadata_file_age_ranges.allLines_total_forPercentage.delivery <- left_join(metadata_file_age_ranges.allLines.delivery,  metadata_file_total_forPercentage_ageRanges)
#divide by total_forPercentage to get prevalence in %
metadata_file_age_ranges.allLines_total_forPercentage.delivery <- metadata_file_age_ranges.allLines_total_forPercentage.delivery %>% 
  mutate(percentage = occurrences*100/total_forPercentage)


metadata_file_age_ranges.allLines_total_forPercentage.delivery <- metadata_file_age_ranges.allLines_total_forPercentage.delivery %>% 
  mutate(Other_variable = fct_relevel(Other_variable, "Total", "Vaginal delivery", "C-section delivery"),
         point_size = case_when(
             (total_forPercentage>=5 & total_forPercentage<50) ~ 1, 
             (total_forPercentage>=50 & total_forPercentage<100) ~ 2,
             (total_forPercentage>=100 & total_forPercentage<500) ~ 3,
             (total_forPercentage>=500) ~ 4,
             #less than n=5 observations supporting the group
             TRUE ~ 0))

readable_age_label <- c("0-1 mos","1-2 mos","2-4 mos","4-6 mos","6-8 mos","8-10 mos","10-12 mos","1-4 yrs","4-7 yrs")

p.delivery<- ggplot(data=metadata_file_age_ranges.allLines_total_forPercentage.delivery, aes(x=factor(age_plot_range, levels = c("0_1M","1_2M","2_4M","4_6M","6_8M","8_10M","10_12M","12_48M","48_84M","84_120M")), y=percentage, group=Other_variable)) +
  geom_line(aes(color=Other_variable, linetype=Other_variable), size = 1.2) +
  scale_linetype_manual(values=c("solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid"))+
  scale_color_manual(values=myPalette) +
  geom_point(pch = 21, color="black", stroke = 1, aes(fill=Other_variable, size = as.factor(point_size))) + 
  scale_fill_manual(values=myPalette) +
  scale_size_discrete(labels=c("100 \u2264 N. samples < 500", "N. samples \u2265  500")) +
  labs(x="Age groups", y = "Prevalence %", size = 10)+
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust = 1.0)) +
  scale_x_discrete(labels= readable_age_label) +
  theme(legend.title=element_blank(),
  legend.position = "none")

p.delivery
```

#plot totals line with prematurity line
```{r}
metadata_file_age_ranges.allLines.prematurity <- bind_rows(metadata_file_age_ranges.all, metadata_file_age_ranges.Premature)

# total_forPercentage provides the total number of samples available oer age range without specific filters (i.e. both cdiff+ and cdiff- timeseries are included)
metadata_file_age_ranges.allLines_total_forPercentage.prematurity <- left_join(metadata_file_age_ranges.allLines.prematurity,  metadata_file_total_forPercentage_ageRanges)
#divide by total_forPercentage to get prevalence in %
metadata_file_age_ranges.allLines_total_forPercentage.prematurity <- metadata_file_age_ranges.allLines_total_forPercentage.prematurity %>% 
  mutate(percentage = occurrences*100/total_forPercentage)


metadata_file_age_ranges.allLines_total_forPercentage.prematurity <- metadata_file_age_ranges.allLines_total_forPercentage.prematurity %>% 
  mutate(Other_variable = fct_relevel(Other_variable, "Total", "Full-term", "Pre-term"),
         point_size = case_when(
             (total_forPercentage>=5 & total_forPercentage<50) ~ 1, 
             (total_forPercentage>=50 & total_forPercentage<100) ~ 2,
             (total_forPercentage>=100 & total_forPercentage<500) ~ 3,
             (total_forPercentage>=500) ~ 4,
             #less than n=5 observations supporting the group
             TRUE ~ 0))

readable_age_label <- c("0-1 mos","1-2 mos","2-4 mos","4-6 mos","6-8 mos","8-10 mos","10-12 mos","1-4 yrs","4-7 yrs")

p.prematurity <- ggplot(data=metadata_file_age_ranges.allLines_total_forPercentage.prematurity, aes(x=factor(age_plot_range, levels = c("0_1M","1_2M","2_4M","4_6M","6_8M","8_10M","10_12M","12_48M","48_84M","84_120M")), y=percentage, group=Other_variable)) +
  geom_line(aes(color=Other_variable, linetype=Other_variable), size = 1.2) +
  scale_linetype_manual(values=c("solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid"))+
  scale_color_manual(values=myPalette) +
  geom_point(pch = 21, color="black", stroke = 1, aes(fill=Other_variable, size = as.factor(point_size))) + 
  scale_fill_manual(values=myPalette) +
  scale_size_discrete(labels=c("100 \u2264 N. samples < 500", "N. samples \u2265  500")) +
  labs(x="Age groups", y = "Prevalence %", size = 10)+
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust = 1.0)) +
  scale_x_discrete(labels= readable_age_label)+
  theme(legend.title=element_blank(),
  legend.position = "none")

p.prematurity
```

#plot totals line with prematurity line
```{r}
metadata_file_age_ranges.allLines.diet <- bind_rows(metadata_file_age_ranges.all, metadata_file_age_ranges.Diet)

# total_forPercentage provides the total number of samples available oer age range without specific filters (i.e. both cdiff+ and cdiff- timeseries are included)
metadata_file_age_ranges.allLines_total_forPercentage.diet <- left_join(metadata_file_age_ranges.allLines.diet,  metadata_file_total_forPercentage_ageRanges)
#divide by total_forPercentage to get prevalence in %
metadata_file_age_ranges.allLines_total_forPercentage.diet <- metadata_file_age_ranges.allLines_total_forPercentage.diet %>% 
  mutate(percentage = occurrences*100/total_forPercentage)


metadata_file_age_ranges.allLines_total_forPercentage.diet <- metadata_file_age_ranges.allLines_total_forPercentage.diet %>% 
  mutate(Other_variable = fct_relevel(Other_variable, "Total", "mixed_bottlefed_breastfed", "exclusively bottlefed", "exclusively breastfed"),
         point_size = case_when(
             (total_forPercentage>=5 & total_forPercentage<50) ~ 1, 
             (total_forPercentage>=50 & total_forPercentage<100) ~ 2,
             (total_forPercentage>=100 & total_forPercentage<500) ~ 3,
             (total_forPercentage>=500) ~ 4,
             #less than n=5 observations supporting the group
             TRUE ~ 0))

readable_age_label <- c("0-1 mos","1-2 mos","2-4 mos","4-6 mos","6-8 mos","8-10 mos","10-12 mos","1-4 yrs","4-7 yrs")

p.diet <- ggplot(data=metadata_file_age_ranges.allLines_total_forPercentage.diet, aes(x=factor(age_plot_range, levels = c("0_1M","1_2M","2_4M","4_6M","6_8M","8_10M","10_12M","12_48M","48_84M","84_120M")), y=percentage, group=Other_variable)) +
  geom_line(aes(color=Other_variable, linetype=Other_variable), size = 1.2) +
  scale_linetype_manual(values=c("solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid", "solid"))+
  scale_color_manual(values=myPalette) +
  geom_point(pch = 21, color="black", stroke = 1, aes(fill=Other_variable, size = as.factor(point_size))) + 
  scale_fill_manual(values=myPalette) +
  scale_size_discrete(labels=c("100 \u2264 N. samples < 500", "N. samples \u2265  500")) +
  labs(x="Age groups", y = "Prevalence %", size = 10)+
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust = 1.0)) +
  scale_x_discrete(labels= readable_age_label) +
  theme(legend.title=element_blank(),
       legend.position = "none")

p.diet
```


#arrange and save plots
```{r}
pdf(file="figures/timeseries_appearance/first_appearance_cdiff_allLines.pdf", width=8, height=6)
grid.arrange(p.status, p.delivery, p.prematurity, p.diet, nrow = 2)
dev.off()
```

