---
title: "C. difficile prevalence and abundance analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
```

#prepare environment
```{r}
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(gridExtra)
library(dplyr)
library("reshape2")
library("grid")

size_text = 10
```

#load and filter metadata table 
```{r}
#after timeseries dereplication: one sample per subject or per timeseries
metadata_file <- read_csv("data/metadata_timeseries_dereplicated.csv", 
                          col_types = cols_only(
                            Samples = col_character(),
                            Study = col_factor(), 
                            Environment = col_factor(), 
                            Gut_sample_type = col_factor(),
                            Host = col_factor(),
                            Host_species = col_factor(),
                            Animal_age = col_factor(),
                            Age_group = col_factor(), 
                            Age_months = col_character(), 
                            Status = col_factor(), 
                            Premature = col_factor(),
                            Delivery_mode = col_factor(),
                            Geography = col_factor(), 
                            Continent = col_factor(), 
                            CDI = col_factor(), 
                            Cdiff = col_factor(),
                            final_representative = col_character(), 
                            diet = col_character()
))

#preliminary filtering
metadata_file2 <- metadata_file %>%
    filter(Host=="Human" & Environment=="Gut" & Gut_sample_type=="Stool" & final_representative=="Yes" & !is.na(Age_months))
```

#animal samples
```{r}
metadata_file_animal <- metadata_file %>%
    filter(Host=="Animal" & Environment=="Gut" & Gut_sample_type=="Stool" & final_representative=="Yes", 
           Study != "Grzeskowiak_2017_cdiff_pig") 

total <- metadata_file_animal %>% 
  group_by(Host_species) %>%
  summarise(count = n())

cdiffPos <- metadata_file_animal %>%
  filter(Cdiff == "Yes") %>% 
  group_by(Host_species) %>%
  summarise(count_pos = n()) 

#calculate prevalence values
prevalence.plot <- merge(total, cdiffPos, all = TRUE) %>% 
  mutate(prevalence = count_pos*100/count) %>% 
  #keep species with more than 10 samples available
  filter(count >10)

prevalence.plot[is.na(prevalence.plot)] = 0

readable_species_label <- c("Dog", "Cat", "Cow", "Pig", "Vulture", "Chicken")

prevalence.plot.sorted <- prevalence.plot %>% 
  filter(prevalence != 0) %>% 
  mutate(Host_species = fct_relevel(Host_species, "Canis lupus familiaris", "Felis catus", "Bos taurus", "Sus scrofa domesticus", "Gymnogyps californianus" , "Gallus gallus domesticus"))

prevalence.plot.sorted <- prevalence.plot.sorted %>% 
  mutate(point_size = case_when(
             (count>=5 & count<50) ~ 1, 
             (count>=50 & count<100) ~ 2,
             (count>=100 & count<500) ~ 3,
             (count>=500) ~ 3,
             #less than n=5 observations supporting the group
             TRUE ~ 0)) %>% 
  filter(point_size != 0) %>% 
  mutate(Vertebrate_class = case_when(
             (Host_species == "Gallus gallus domesticus" || Host_species == "Gymnogyps californianus") ~ "Bird", 
             TRUE ~ "Mammal"))
         
colors <- c('#e66101','#e66101')

p0 <- ggplot(prevalence.plot.sorted, aes(x=Host_species, y=prevalence, group=Vertebrate_class)) +
  geom_segment( aes(x=Host_species, xend=Host_species, y=0, yend=prevalence), size = 1) +
  geom_point(shape=21, color="black", stroke = 2, aes(fill=Vertebrate_class, size = as.factor(point_size))) +
  scale_fill_manual(values=colors) +
  labs(x="Host species", y = "Prevalence (%)")+
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, vjust=1.0, hjust = 1.0, size = size_text),
        axis.title.x = element_text(size= size_text),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none") +
  guides(fill = FALSE) +
  ylim(0,100)+
  scale_x_discrete(labels= readable_species_label) +
  scale_size_discrete(labels=c("5 \u2264 N. samples < 50", "100 \u2264 N. samples < 500", "N. samples \u2265  500")) +
  ylim(0,100)
```

#define age ranges of interest
```{r}
infant_interval = 2 #months
child_adol_interval = 12*3 #3yrs
adult_elderly_interval = 12*10 #10yrs

age_range_as_label <- function(age_range) {
  return(paste0(age_range[1], "_", age_range[2], "M"))
}

replace_age_range <- function(age_ranges, i, sequence, interval) {
  age_ranges[[length(age_ranges)+ 1]] <- c(i, i+interval)
  return(age_ranges)
}

age_ranges <- list()
#ad hoc age ranges 0-1M and 1-2M
age_ranges[[length(age_ranges)+ 1]] <- c(0, 1)
age_ranges[[length(age_ranges)+ 1]] <- c(1, 2)

#infants age ranges defined by equal time intervals
seq_infant <- seq(2, 12-infant_interval, by=infant_interval)
for (i in seq_infant){
  age_ranges <- replace_age_range(age_ranges, i, seq_infant, infant_interval)
}

#children and adolescents
seq_child_adol <- seq(12, 228-child_adol_interval, by=child_adol_interval)
for (i in seq_child_adol){
  age_ranges <- replace_age_range(age_ranges, i, seq_child_adol, child_adol_interval)
}

#adults and elderly
seq_adult_elderly <- seq(228, 1308-adult_elderly_interval, by=adult_elderly_interval)
for (i in seq_adult_elderly){
  age_ranges <- replace_age_range(age_ranges, i, seq_adult_elderly, adult_elderly_interval)
}

age_ranges_labels <- mapply(age_range_as_label, age_ranges)
```

#manipulate metadata table
```{r}
convert_range <- function(val, age_ranges) {
  values <- str_split(val, "-")
  # If we have an age range
  if(length(values) == 2) {
    values <- as.double(values)
    for (age_range in age_ranges){
      if (values[1] >= age_range[1] && values[1] < age_range[2] && 
          values[2] >= age_range[1] && values[2] < age_range[2]) {
        return(age_range_as_label(age_range))
      }
    }
  }
  
  # if the age range didn't fit into any category it returns the age interval itself
  if(str_detect(val, "-")) {return(val)}
  
  # Assume that anything at this point is a number
  dval <- as.double(val)
  # If the value is not a number return whatever was there to begin with
  if(is.na(dval)) {stop(paste0("Unexpected value in specified column (Age_months?)", " '", val, "'"))}
  # from this point onwards only numbers should remain
  for (age_range in age_ranges){
    if (dval >= age_range[1] && dval < age_range[2]) {
      return(age_range_as_label(age_range))
    }
  }
  return(dval)
}

convert_to_ranges <- function(age_months, age_ranges){
  mapply(convert_range, age_months, MoreArgs=list(age_ranges=age_ranges))
}

metadata_file_age_ranges <- metadata_file2 %>% 
  mutate(age_plot_range = convert_to_ranges(Age_months, age_ranges),
         age_plot_range = case_when(
           #merge in >79yrs
           (age_plot_range == "948_1068M") ~ ">948M", 
           (age_plot_range == "1068_1188M") ~ ">948M",
           (age_plot_range == "1188_1308M") ~ ">948M",
           #merge 10-19 yrs
           (age_plot_range == "120_156M") ~ "120_228M",
           (age_plot_range == "156_192M") ~ "120_228M",
           (age_plot_range == "192_228M") ~ "120_228M",
           #keep other age categories unchanged
           TRUE ~ age_plot_range))

metadata_file_age_ranges.tmp <- metadata_file_age_ranges %>% 
  filter(Host=="Human" & Environment=="Gut" & Gut_sample_type=="Stool" & final_representative=="Yes") %>% 
  mutate(cdiff_bool = ifelse(Cdiff=="Yes", TRUE, FALSE))
```

#Get prevalence data divided by health status
```{r}
get_prevalence <- function(data.input, status) {
    data.input %>%
    filter(Status == status & str_detect(age_plot_range, "-") == FALSE) %>% 
    group_by(age_plot_range) %>% 
    dplyr::summarise(total = n(), cdiffPos = sum(cdiff_bool)) %>% 
    mutate(percentages = ifelse(is.na(cdiffPos), 0, 100*cdiffPos/total), Status = status)
}

partial_healthy <- get_prevalence(metadata_file_age_ranges.tmp, "Healthy")
partial_diseased <- get_prevalence(metadata_file_age_ranges.tmp, "Diseased")
table.plot.status <- rbind(partial_healthy, partial_diseased)

table.plot.status <- table.plot.status %>% 
  mutate(point_size = case_when(
             (total>=5 & total<50) ~ 1, 
             (total>=50 & total<100) ~ 2,
             (total>=100 & total<500) ~ 3,
             (total>=500) ~ 3,
             #less than n=5 observations supporting the group
             TRUE ~ 0))
```

#Plot prevalence data divided by status
```{r}

# kicks out data groups supported by less the n=5 observations (of total samples)
table.plot.status <- table.plot.status %>% 
  filter(point_size != 0)

colors <- c('#5e3c99','#e66101')

readable_age_label <- c("0-1 mos","1-2 mos","2-4 mos","4-6 mos","6-8 mos","8-10 mos","10-12 mos","1-4 yrs","4-7 yrs","7-10 yrs","10-19 yrs","19-29 yrs","29-39 yrs","39-49 yrs","49-59 yrs","59-69 yrs","69-79 yrs",">79 yrs")

p1 <- ggplot(data=table.plot.status, aes(x=factor(age_plot_range, levels = c("0_1M","1_2M","2_4M","4_6M","6_8M","8_10M","10_12M","12_48M","48_84M","84_120M","120_228M","228_348M","348_468M","468_588M","588_708M","708_828M","828_948M",">948M")), y=percentages, group=Status)) +
  geom_line(aes(color=Status), size = 2) +
  scale_color_manual(values=colors) +
  geom_point(pch = 21, color="black", stroke = 2, aes(fill=Status, size = as.factor(point_size))) + 
  scale_fill_manual(values=colors) +
  scale_size_discrete(labels=c("5 \u2264 N. samples < 50", "50 \u2264 N. samples < 100", "100 \u2264 N. samples < 500", "N. samples \u2265  500")) +
  scale_x_discrete(labels= readable_age_label) +
  labs(x="Age groups", y = "Prevalence (%)")+
  theme_bw() +
  theme(plot.title = element_text(family = "Helvetica", face = "bold", size = size_text), 
       axis.text.x = element_text(angle=45, vjust=1.0, hjust = 1.0),
       axis.title.x = element_text(size= size_text),
      axis.title.y = element_text(size= size_text)) +
  guides(fill=FALSE) +
  theme(legend.title = element_blank(),
  legend.text = element_text(size = size_text)) +
  theme(
    legend.position = c(.95, .95),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6)
    ) +
  ylim(0,100)

p1

```

#Get prevalence data divided by health status and delivery mode
```{r}
metadata_file_age_ranges.tmp.pooled <- metadata_file_age_ranges.tmp %>% 
  mutate(age_plot_range = case_when(
           #merge first semester
           (age_plot_range == "1_2M") ~ "1_6M", 
           (age_plot_range == "2_4M") ~ "1_6M",
           (age_plot_range == "4_6M") ~ "1_6M",
           #merge second semester
           (age_plot_range == "6_8M") ~ "6_12M",
           (age_plot_range == "8_10M") ~ "6_12M",
           (age_plot_range == "10_12M") ~ "6_12M",
           #keep other age categories unchanged
           TRUE ~ age_plot_range))

get_prevalence <- function(data.input, status, deliveryMode) {
    data.input %>%
    filter(Delivery_mode == deliveryMode & Status == status & str_detect(age_plot_range, "-") == FALSE & (Age_group == "Infant" | Age_group == "Child")) %>% 
    group_by(age_plot_range) %>% 
    dplyr::summarise(total = n(), cdiffPos = sum(cdiff_bool)) %>% 
    mutate(percentages = ifelse(is.na(cdiffPos), 0, 100*cdiffPos/total), status_delivery = paste0(status," ", deliveryMode))
}

partial_healthy_vaginal <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Healthy", "Vaginal")
partial_healthy_cesarean <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Healthy", "Cesarean")
partial_diseased_vaginal <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Diseased", "Vaginal")
partial_diseased_cesarean <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Diseased", "Cesarean")
table.plot.delivery <- rbind(partial_healthy_vaginal, partial_healthy_cesarean, partial_diseased_vaginal, partial_diseased_cesarean)

to_not_plot <- c("48_84M", "84_120M","120_156M","156_192M","192_228M","228_348M","348_468M","468_588M","588_708M","708_828M","828_948M","948_1068M","1068_1188M","1188_1308M")
`%!in%` = Negate(`%in%`)

table.plot.delivery <- table.plot.delivery %>%
  filter(age_plot_range %!in% to_not_plot) %>% 
  #calculate percentage of prevalence
  mutate(point_size = case_when(
             (total>=5 & total<50) ~ 1, 
             (total>=50 & total<100) ~ 2,
             (total>=100 & total<500) ~ 3,
             (total>=500) ~ 3,
             #less than n=5 observations supporting the group
             TRUE ~ 0))
```


#Plot prevalence data divided by health status and delivery mode 
```{r}
colors <- c('#5e3c99', '#b2abd2', '#fdb863', '#e66101')
readable_age_label <- c("0-1 mos","1-6 mos","6-12 mos","1-4 yrs")

table.plot.delivery <- table.plot.delivery %>% 
  filter(point_size != 0)

p2 <- ggplot(data=table.plot.delivery, aes(x=factor(age_plot_range, levels = c("0_1M","1_6M","6_12M", "12_48M")), y=percentages, group=status_delivery)) +
  geom_line(aes(color=status_delivery), size = 2) +
  scale_color_manual(values=colors) +
  geom_point(pch = 21, color="black", stroke = 2, aes(fill=status_delivery, size = as.factor(point_size)), show.legend = FALSE) + 
  scale_fill_manual(values=colors) +
  labs(x="Age groups", y = "Prevalence (%)")+
  theme_bw() +
  guides(fill=FALSE) +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = size_text),
        axis.title.x = element_text(size= size_text),
        axis.title.y = element_text(size= size_text)) +
  ylim(0,100) +
  scale_x_discrete(labels= readable_age_label) +
  theme(
    legend.position = c(.95, .25),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6)
    )

p2
```


#Get prevalence data divided by health status and prematurity
```{r}
get_prevalence <- function(data.input, status, prematurity) {
    data.input %>%
    filter(Premature == prematurity & Status == status & str_detect(age_plot_range, "-") == FALSE & (Age_group == "Infant" | Age_group == "Child")) %>% 
    group_by(age_plot_range) %>% 
    dplyr::summarise(total = n(), cdiffPos = sum(cdiff_bool)) %>% 
    mutate(percentages = ifelse(is.na(cdiffPos), 0, 100*cdiffPos/total), status_prematurity = paste0(status,"_Premature", prematurity))
}

partial_healthy_premature <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Healthy", "Yes")
partial_healthy_atTerm <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Healthy", "No")
partial_diseased_premature <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Diseased", "Yes")
partial_diseased_atTerm <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Diseased", "No")
table.plot.prematurity <- rbind(partial_healthy_premature, partial_healthy_atTerm, partial_diseased_premature, partial_diseased_atTerm)

to_not_plot <- c("48_84M", "84_120M","120_156M","156_192M","192_228M","228_348M","348_468M","468_588M","588_708M","708_828M","828_948M","948_1068M","1068_1188M","1188_1308M")
`%!in%` = Negate(`%in%`)

table.plot.prematurity <- table.plot.prematurity %>%
  filter(age_plot_range %!in% to_not_plot) %>% 
  #calculate percentage of prevalence
  mutate(point_size = case_when(
             (total>=5 & total<50) ~ 1, 
             (total>=50 & total<100) ~ 2,
             (total>=100 & total<500) ~ 3,
             (total>=500) ~ 3,
             #less than n=5 observations supporting the group
             TRUE ~ 0),
         prematurity_labels = case_when(
             (status_prematurity=="Healthy_PrematureYes") ~ "Healthy pre-term", 
             (status_prematurity=="Healthy_PrematureNo") ~ "Healthy full-term", 
             (status_prematurity=="Diseased_PrematureYes") ~ "Diseased pre-term", 
             TRUE ~ "Diseased full-term"),
         prematurity_labels = fct_relevel(prematurity_labels, 
                                          "Diseased full-term", "Diseased pre-term", "Healthy pre-term", "Healthy full-term"))
```


#Plot prevalence data divided by health status and prematurity 
```{r}
readable_age_label <- c("0-1 mos","1-6 mos","6-12 mos","1-4 yrs")

table.plot.prematurity <- table.plot.prematurity %>% 
  filter(point_size != 0)

p3 <- ggplot(data=table.plot.prematurity, aes(x=factor(age_plot_range, levels = c("0_1M","1_6M","6_12M", "12_48M")), y=percentages, group=prematurity_labels)) +
  geom_line(aes(color=prematurity_labels), size = 2) +
  scale_color_manual(values=colors) +
  geom_point(pch = 21, color="black", stroke = 2, aes(fill=prematurity_labels, size = as.factor(point_size)), show.legend = FALSE) + 
  scale_fill_manual(values=colors) +
  labs(x="Age groups", y = "Prevalence (%)")+
  theme_bw() +
  guides(fill=FALSE) +
  theme(legend.title = element_blank(), 
        axis.title.x = element_text(size= size_text),
        axis.title.y = element_text(size= size_text)) +
  ylim(0,100) +
  scale_x_discrete(labels= readable_age_label) +
  theme(
    legend.position = c(.95, .25),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6)
    )
p3
```

#Get prevalence data divided by feeding mode
```{r}
get_prevalence <- function(data.input, feedingMode) {
    data.input %>%
    filter(diet == feedingMode & str_detect(age_plot_range, "-") == FALSE & (Age_group == "Infant" | Age_group == "Child")) %>% 
    filter(!is.na(diet), age_plot_range != "6_8M", age_plot_range != "8_10M", age_plot_range != "10_12M", age_plot_range != "12_48M") %>%
    mutate(age_plot_range = case_when(
      #only 0-1 and 1-6 months
      (age_plot_range == "0_1M") ~ "0_1M", 
      TRUE ~ "1_6M")) %>% 
    group_by(age_plot_range) %>% 
    dplyr::summarise(total = n(), cdiffPos = sum(cdiff_bool)) %>% 
    mutate(percentages = ifelse(is.na(cdiffPos), 0, 100*cdiffPos/total), diet = feedingMode)
}

partial_breastFed <- get_prevalence(metadata_file_age_ranges.tmp, "exclusively breastfed")
partial_bottle <- get_prevalence(metadata_file_age_ranges.tmp, "exclusively bottlefed")
partial_mixed <- get_prevalence(metadata_file_age_ranges.tmp, "mixed_bottlefed_breastfed")
table.plot <- rbind(partial_breastFed, partial_bottle, partial_mixed)

#calculate prevalence percentage
table.plot.diet <- table.plot %>% 
  mutate(point_size = case_when(
             (total>=5 & total<50) ~ 1, 
             (total>=50 & total<100) ~ 2,
             (total>=100 & total<500) ~ 3,
             (total>=500) ~ 3,
             #less than n=5 observations supporting the group
             TRUE ~ 0),
  diet_labels = case_when(
             (diet=="exclusively breastfed") ~ "Exclusive breast milk", 
             (diet=="exclusively bottlefed") ~ "Exclusive formula", 
             TRUE ~ "Mixed"))
```


#Plot prevalence data divided by feeding mode
```{r}
colors <- c('#b2df8a', '#1f78b4', '#a6cee3')
readable_age_label <- c("0-1 mos","2-6 mos")

table.plot.diet <- table.plot.diet %>% 
  filter(point_size != 0)

p4 <- ggplot(data=table.plot.diet, aes(x=age_plot_range, y=percentages, group=diet_labels)) +
  geom_line(aes(color=diet_labels), size = 2) +
  scale_color_manual(values=colors) +
  geom_point(pch = 21, color="black", stroke = 2, aes(fill=diet_labels, size = as.factor(point_size)), show.legend = FALSE) + 
  scale_fill_manual(values=colors) +
  labs(x="Age groups", y = "Prevalence (%)")+
  theme_bw() +
  theme(axis.text.x = element_text(angle=0)) +
  guides(fill=FALSE) +
  theme(legend.title = element_blank(),
        axis.title.x = element_text(size= size_text),
        axis.title.y = element_text(size= size_text)) +
  ylim(0,100) +
  scale_x_discrete(labels= readable_age_label) +
  theme(
    legend.position = c(.95, .25),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6)
    )
p4
```

#Plot prevalence data divided by health, gestational age AND delivery mode
```{r}
colors <- c( '#1f78b4','navy', 'aquamarine4', 'aquamarine1', 'goldenrod1', 'darkorange', "darkred", "black")
readable_age_label <- c("0-1 mos","1-6 mos","6-12 mos","1-4 yrs")

get_prevalence <- function(data.input, status, prematurity, deliveryMode) {
    data.input %>%
    filter(Status == status & Delivery_mode == deliveryMode & Premature == prematurity & str_detect(age_plot_range, "-") == FALSE & (Age_group == "Infant" |Age_group == "Child")) %>% 
    group_by(age_plot_range) %>% 
    dplyr::summarise(total = n(), cdiffPos = sum(cdiff_bool)) %>% 
    mutate(prematurity_expanded = ifelse(prematurity=="Yes", "Pre-term","full-term")) %>% 
    mutate(percentages = ifelse(is.na(cdiffPos), 0, 100*cdiffPos/total), term_mode = paste0(status, "_", prematurity_expanded,"_", deliveryMode))
}

partial_hfullterm_vag <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Healthy", "No", "Vaginal" )
partial_hfullterm_csec <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Healthy", "No", "Cesarean")
partial_hpreterm_vag <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Healthy", "Yes", "Vaginal")
partial_hpreterm_csec <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Healthy", "Yes", "Cesarean")
partial_dfullterm_vag <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Diseased", "No", "Vaginal" )
partial_dfullterm_csec <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Diseased", "No", "Cesarean")
partial_dpreterm_vag <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Diseased", "Yes", "Vaginal")
partial_dpreterm_csec <- get_prevalence(metadata_file_age_ranges.tmp.pooled, "Diseased", "Yes", "Cesarean")

table.plot.term.deliveryMode <- rbind(partial_hfullterm_vag, partial_hfullterm_csec, partial_hpreterm_vag, partial_hpreterm_csec,
                                      partial_dfullterm_vag, partial_dfullterm_csec, partial_dpreterm_vag, partial_dpreterm_csec)

table.plot.term.deliveryMode <- table.plot.term.deliveryMode %>% 
  #calculate percentage of prevalence
  mutate(point_size = case_when(
             (total>=5 & total<50) ~ 1, 
             (total>=50 & total<100) ~ 2,
             (total>=100 & total<500) ~ 3,
             (total>=500) ~ 3,
             #less than n=5 observations supporting the group
             TRUE ~ 0))

table.plot.term.deliveryMode <- table.plot.term.deliveryMode %>% 
  filter(point_size != 0)

combinations <- ggplot(data=table.plot.term.deliveryMode, aes(x=factor(age_plot_range, levels = c("0_1M","1_6M","6_12M", "12_48M")), y=percentages, group=term_mode)) +
  geom_line(aes(color=term_mode), size = 2) +
  scale_color_manual(values=colors) +
  geom_point(pch = 21, color="black", stroke = 2, aes(fill=term_mode, size = as.factor(point_size)), show.legend = FALSE) + 
  scale_fill_manual(values=colors) +
  labs(x="Age groups", y = "Prevalence (%)")+
  theme_bw() +
  theme(axis.text.x = element_text(angle=0)) +
  guides(fill=FALSE) +
  theme(legend.title = element_blank(),
        axis.title.x = element_text(size= size_text),
        axis.title.y = element_text(size= size_text)) +
  ylim(0,100) 

pdf(file="figures/prevalence_Cdiff/prevalence_combinations.pdf", width=10, height=6)
combinations
dev.off()

```

#grouped barplot for geography (latest version)
```{r}

map.tmp <- metadata_file %>%
  filter(Host=="Human" & Environment=="Gut" & Gut_sample_type=="Stool" & final_representative=="Yes" & Geography != "NA") %>%
  mutate(cdiff_bool = ifelse(Cdiff=="Yes", TRUE, FALSE))

map.tmp.all <- map.tmp %>%
  filter((Age_group == "Infant" | Age_group == "Adult") & Status == "Healthy") %>%
  group_by(Geography, Continent, Age_group) %>%
  mutate(Age_group= fct_relevel(Age_group, "Infant", "Adult")) %>% 
  dplyr::summarise(total = n(), cdiffPos = sum(cdiff_bool)) %>%
  mutate(percentages = ifelse(is.na(cdiffPos), 0, 100*cdiffPos/total)) %>% 
  filter(as.numeric(percentages)>1) %>% 
  ungroup() %>% 
  group_by(Continent, Age_group) %>%
  mutate(sum_continent= sum(percentages)) %>% 
  ungroup()

means <- map.tmp.all %>% group_by(Continent) %>% summarise(mean=mean(percentages))

geo_infants <- ggplot(data=map.tmp.all %>% filter(Age_group=="Infant"), 
       aes(x=reorder(Geography, -sum_continent), y=percentages, fill=Continent)) +
  geom_bar(stat="identity", color="#444444") +
  geom_text(aes(label=total), vjust=-0.3, color="black", size=3, position = "dodge") +
  geom_text(aes(label=paste0(round(percentages, digits = 1), "%")), 
            vjust=1.4, color="black", size=3, position = "dodge") +
  labs(y = "Prevalence C. difficile (%)")+
  theme_bw() +
  theme(text = element_text(size=12),
        axis.title.x=element_blank()) +
  ylim(0,100)+
  theme(legend.position = "none")
geo_infants

geo_adults <- ggplot(data=map.tmp.all %>% filter(Age_group=="Adult"), 
       aes(x=reorder(Geography, -sum_continent), y=percentages, fill=Continent)) +
  geom_bar(stat="identity", color="#444444") +
  geom_text(aes(label=total), vjust=-0.3, color="black", size=3, position = "dodge") +
  geom_text(aes(label=paste0(round(percentages, digits = 1), "%")), 
            vjust=1.4, color="black", size=3, position = "dodge") +
  labs(y = "Prevalence C. difficile (%)")+
  theme_bw() +
  theme(text = element_text(size=12),
        axis.title.x=element_blank()) +
  ylim(0,100)+
  theme(legend.position = "none") 

pdf(file="figures/prevalence_Cdiff/prevalence_geography.pdf", width=11, height=4)
grid.arrange(geo_infants, geo_adults, widths = c(0.5, 0.18), heights = c(1.0, 0.1))
dev.off()


```


#create merged panel of plots
```{r}
pdf(file="figures/prevalence_Cdiff/figure_4.pdf", width=10, height=8)
grid.arrange(p0, p1,                                    # bar plot spanning two columns
             p2, p3, p4,                              # box plot and scatter plot
             ncol = 5, nrow = 2, 
             layout_matrix = rbind(c(1,1,1,1,0), c(2,2,3,3,4)))
dev.off()
```
