---
title: "Random mixed effect models for CDI"
author: "Jakob Wirbel"
date: "2021-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## Libraries

```{r load_libraries, message=FALSE}
library("tidyverse")
library("here")
library("lmerTest")
library("ggrepel")
library("ComplexHeatmap")
```

## Data

First, we load all the data:
```{r load data, message=FALSE}
# load data
path_file <- "data/CDI/"
meta.all <- read_csv('data/CDI/CDI_mergedControl_samples.csv')
feat.all <- read.table('data/CDI/mOTU20_profiles_filteredCDI.csv',
                       sep=',', stringsAsFactors = FALSE, check.names = FALSE,
                       comment.char = '', quote = '"', header = TRUE, 
                       row.names = 1)
feat.rel <- prop.table(as.matrix(feat.all), 2)
```

We need to adjust a few columns in the metadata:
```{r adjust_meta}
meta.all <- meta.all %>% 
  # include FMT samples as CDI-positive?
  mutate(CDI=case_when(is.na(CDI)~'Yes', TRUE~CDI)) %>% 
  # include sequencing depth (as measured by mOTU counts) as new column
  left_join(enframe(colSums(feat.all), name='Samples', value='Depth'), 
            by='Samples') %>% 
  mutate(Depth=as.factor(
    cut(Depth, breaks = quantile(Depth), labels=c(1,2,3,4))))
```

We also need to filter the feature table somehow. Here, we filter by prevalence
per study:
```{r prev_filter}
prev.mat <- do.call(cbind, map(unique(meta.all$Study), .f = function(x){
  # ignore studies with fewer than 10 samples
  if ((meta.all %>% filter(Study==x) %>% nrow) < 10){
    return(rep(NA_real_, nrow(feat.rel)))
    }
  return(rowMeans(feat.rel[,meta.all %>% 
                             filter(Study==x) %>% 
                             pull(Samples)] > 0))
  }))
colnames(prev.mat) <- unique(meta.all$Study)
prev.mat <- prev.mat[,colSums(is.na(prev.mat))!=nrow(prev.mat)]

# filter criterion: more than 5% prevalence in 3 or more studies
f.idx <- which(rowSums(prev.mat > 0.05) >=3)
feat.rel.filt <- feat.rel[names(f.idx),meta.all$Samples]
```

# Confounder analysis

Similar analysis to what we did in the CRC meta-analysis:
```{r disease_effect}
ss.disease <- apply(feat.rel.filt, 1, FUN=function(x, label){
  rank.x <- rank(x)/length(x)
  ss.tot <- sum((rank.x - mean(rank.x))^2)/length(rank.x)
  ss.o.i <- sum(vapply(unique(label), function(l){
    sum((rank.x[label==l] - mean(rank.x[label==l]))^2)
  }, FUN.VALUE = double(1)))/length(rank.x)
  return(1-ss.o.i/ss.tot)
}, label=meta.all %>% pull(CDI))
```

Now, we can prepare the result matrix:
```{r pre_conf}
t.mean <- apply(feat.rel.filt, 1, mean, trim=0.1)

df.plot.all <- tibble(
  species=rownames(feat.rel.filt),
  disease=ss.disease,
  t.mean=t.mean)
```

Finally, we can calculate the effect for different confounding factors:
```{r calc_conf, message=FALSE}
for (meta.var in c('Study', 'Age_group', 'Sex', 'Abx', "Depth", "Status")){
  
  message('# ********************************\n# ', meta.var)
  meta.c <- meta.all %>%
    filter(!is.na(eval(parse(text=meta.var))))
  
  # message('# After filtering, the distribution of variables is:\n')
  # print(table(meta.c$CDI, meta.c %>% pull(meta.var)))
  # print(table(meta.c$Study))
  feat.red <- feat.rel.filt[,meta.c$Samples]
  
  message('# Calculating variance explained by meta-variable...')
  ss.var <- apply(feat.red, 1, FUN=function(x, label){
    rank.x <- rank(x)/length(x)
    ss.tot <- sum((rank.x - mean(rank.x))^2)/length(rank.x)
    ss.o.i <- sum(vapply(unique(label), function(l){
      sum((rank.x[label==l] - mean(rank.x[label==l]))^2)
    }, FUN.VALUE = double(1)))/length(rank.x)
    return(1 - ss.o.i/ss.tot)
  }, label=meta.c %>% pull(meta.var))
  df.plot.all[[meta.var]] <- ss.var
  
  message('# Calculating association with the meta-variable...')
  if (meta.c %>% pull(meta.var) %>% unique %>% length > 2){
    meta.significance <- apply(feat.red, 1, FUN=function(x, var){
      kruskal.test(x~as.factor(var))$p.value
    }, var=meta.c %>% pull(meta.var))
  } else {
    meta.significance <- apply(feat.red, 1, FUN=function(x, var){
      wilcox.test(x~as.factor(var))$p.value
    }, var=meta.c %>% pull(meta.var))
  }
  meta.significance <- p.adjust(meta.significance, method='fdr')
  df.plot.all[[paste0(meta.var, '.significance')]] <- meta.significance
}

```

Finally, plot everything together:
```{r plot_conf}
df.plot.all %>%
  pivot_longer(-c(species, disease,t.mean), 
               names_to='type', values_to='meta') %>%
  filter(!str_detect(type, '.significance')) %>%
  filter(complete.cases(.)) %>% 
  ggplot(aes(x=disease, y=meta, size=t.mean+1e-08)) +
    geom_abline(slope = 1, intercept = 0, col='darkgrey', lty=3) + 
    geom_point(shape=19, alpha=.5) +
    xlab('Variance explained by disease status') +
    ylab('Variance explained by metadata variable') +
    theme_bw() +
    facet_wrap(~type, ncol=3) +
    theme(strip.background = element_blank(),
          panel.grid.minor = element_blank()) +
    scale_size_area(name='Trimmed mean\nabundance',
                    breaks=c(1e-07, 1e-06, 1e-05)) +
    guides( size = "legend", colour='legend')
```

The biggest effects seem to come from `Age_group` and `Study`, so we will 
include these in the mixed effect model.

# REM-testing

## With Cdiff-positive samples

For the mixed random effect models, we first need to adjust the metadata again:

```{r adjust_meta_2}
# testing with linear mixed model
meta.all <- meta.all %>% 
  mutate(Age_group=case_when(is.na(Age_group)~'NA', TRUE~Age_group)) %>%
  mutate(Sex=case_when(is.na(Sex)~'NA', TRUE~Sex)) %>%
  mutate(Abx=case_when(is.na(Abx)~'NA', TRUE~Abx))
```

Now, we can run a test with the following model:
> model: feat ~ CDI | Study + Age_group


```{r testing, message=FALSE, warning=FALSE}
df.res <- tibble(p.val=double(0),
                 effect.size=double(0),
                 species=character(0))

for (x in rownames(feat.rel.filt)){
  tmp <- log10(feat.rel.filt[x,meta.all$Samples] + 1e-05)

  df <- data.frame(value=tmp, CDI=meta.all$CDI, Study=meta.all$Study,
                   Age_group=meta.all$Age_group, Abx=meta.all$Abx,
                   Sex=meta.all$Sex)
  res <- lmer(value~CDI + (1|Study) + (1|Age_group), 
              data=df)
  res2 <- summary(res)$coefficients
  df.res <- df.res %>% 
    add_row(p.val=res2[2,5], 
            effect.size=res2[2,1],
            species=x)
}

# save the results
df.res <- df.res %>% 
  filter(species!='-1') %>% 
  # adjust the p-value for multiple testing
  mutate(p.adj=p.adjust(p.val, method='fdr'))
write_tsv(df.res, 'data/CDI/random_effect_model_results.tsv')
```

## Without Cdiff-positive samples

Do the same again, but filter out samples that have detectable levels of 
`Cdiff`:
```{r test_without_cdiff, message=FALSE, warning=FALSE}
df.res.2 <- tibble(p.val=double(0),
                   effect.size=double(0),
                   species=character(0))
meta.all.red <- meta.all %>% 
  filter(Cdiff=='No')

for (x in rownames(feat.rel.filt)){
  tmp <- log10(feat.rel.filt[x,meta.all.red$Samples] + 1e-06)
  
  df <- data.frame(value=tmp, CDI=meta.all.red$CDI, Study=meta.all.red$Study,
                   Age_group=meta.all.red$Age_group, Abx=meta.all.red$Abx,
                   Sex=meta.all.red$Sex)
  res <- lmer(value~CDI + (1|Study) + (1|Age_group), 
              data=df)
  res2 <- summary(res)$coefficients
  df.res.2 <- df.res.2 %>% 
    add_row(p.val=res2[2,5], 
            effect.size=res2[2,1],
            species=x)
}

# adjust for multiple testing
df.res.2 <- df.res.2 %>% 
  filter(species!='-1') %>% 
  # adjust the p-value for multiple testing
  mutate(p.adj=p.adjust(p.val, method='fdr'))
```

## Compare the results

We can have a look how similar the results are with both approaches. Note that
fewer samples means lower power, so the p-values are expected to be a 
bit lower:
```{r combine}
df.plot <- df.res %>% 
  full_join(df.res.2 %>% transmute(species, p.val.without.cdiff=p.val, 
                                   effect.size.without.cdiff=effect.size,
                                   p.adj.without.cdiff=p.adj),
            by='species')
```

How similar are the effect sizes?

```{r compare_effect_sizes}
df.plot %>% 
  ggplot(aes(x=effect.size, y=effect.size.without.cdiff)) + 
    geom_point() +
    geom_abline(slope = 1, intercept = 0) + 
    geom_point(data=. %>% filter(str_detect(species, 'difficile')), 
               col='red', size=2)
```

How similar are the p-values?

```{r compare_p_vals}
df.plot %>% 
  ggplot(aes(x=-log10(p.val), y=-log10(p.val.without.cdiff))) + 
    geom_point() +
    geom_abline(slope = 1, intercept = 0) + 
    geom_point(data=. %>% filter(str_detect(species, 'difficile')), 
               col='red', size=2)
```

Because of the different scales, it might make sense to compute the Spearman
correlation:
```{r compare_p_vals_spearman}
cor(df.plot$p.val, df.plot$p.val.without.cdiff, method='spearman')
```


# Volcano plot

Based on these data, we can plot a nice volcano plot. First, we add some 
annotations:

```{r prep_volcano}
df.res <- df.res %>% 
  # add groups with information from Panela
  mutate(highlight=case_when(str_detect(species, 'perfringens')~'CDI-similar',
                             str_detect(species, 'innocuum')~'CDI-similar',
                             str_detect(species, 'faecalis')~'CDI-similar',
                             str_detect(species, 'oxytoca')~'CDI-similar',
                             str_detect(species, 'aureus')~'CDI-similar',
                             str_detect(species, 'amalonaticus')~'CDI-similar',
                             str_detect(species, 'aeruginosa')~'CDI-similar',
                             str_detect(species, 'cloacae')~'CDI-similar',
                             str_detect(species, 'difficile')~'Cdiff',
                             effect.size > 0.9 ~ 'enriched-relevant',
                             effect.size > 0.1 & effect.size <= 0.9 & p.adj < 0.01 ~ 'enriched',
                             effect.size < -0.1 & p.adj < 0.01 ~ 'depleted',
                             TRUE~'other')) %>% 
  mutate(label=case_when(highlight%in%c('CDI-similar', 'Cdiff', 'enriched-relevant')~
                           str_remove(species, '\\[(ref|meta)_mOTU_v2_[0-9]{4}\\]'),
                         TRUE~'')) %>% 
  mutate(color_positive=p.adj < 0.01 & effect.size > 0.125) %>% 
  mutate(highlight=factor(highlight, 
                          levels = c('Cdiff', 'CDI-similar', 'enriched', 'enriched-relevant', 
                                     'depleted', 'other')))
```

Then we can create the volcano plot:

```{r volcano, warning=FALSE}
g <- df.res %>% 
  ggplot(aes(x=effect.size, y=-log10(p.adj))) + 
    geom_hline(yintercept = 2, colour='darkgrey', lty=2) + 
    geom_vline(xintercept = c(-0.1, 0.1), colour='darkgrey', lty=2) + 
    geom_point(col='lightgrey') + 
    geom_point(data=df.res %>% filter(highlight!='other'),
               aes(col=highlight)) +
    geom_text_repel(aes(label=label), max.overlaps = 10) + 
    theme_bw() + 
    theme(panel.grid.minor=element_blank()) + 
    xlab('LME model effect size') + 
    ylab('-log10(adj. P-value)') + 
    scale_colour_manual(
      values=c('#E40046', '#8246AF', '#FFA300', '#FFA300', '#307FE2', 'lightgrey'), 
      labels=c('C. difficile', 'CDI-similar', 'CDI-enriched', 'CDI-enriched-relevant',
               'CDI-depleted','other'), name='')

ggsave(g, filename = here("figures/CDI/", 'cdiff_lme_volcano.pdf'),
       width = 10, height = 8,
       useDingbats=FALSE)
print(g)
```

We can also have a look at the distribution of single interesting species:

```{r plot_example_species}
x <- "Clostridioides difficile [ref_mOTU_v2_0051]"
x <- "Enterococcus avium [ref_mOTU_v2_0599]"
x <- "Veillonella parvula [ref_mOTU_v2_1042]"
enframe(feat.rel.filt[x,], name='Samples') %>% 
  right_join(meta.all,
             by='Samples') %>% 
  mutate(value=log10(value + 1e-05)) %>% 
  ggplot(aes(x=Study, y=value, fill=CDI)) + 
    geom_boxplot() + 
    theme(axis.text.x=element_text(angle=45, hjust=1))
```

How would the volcano look like for the testing without `Cdiff`-positive 
samples?

```{r volcano_without_cdiff, warning=FALSE}
g2 <- df.res.2 %>% 
  # add groups with information from Panela
  mutate(highlight=case_when(str_detect(species, 'perfringens')~'CDI-similar',
                             str_detect(species, 'innocuum')~'CDI-similar',
                             str_detect(species, 'faecalis')~'CDI-similar',
                             str_detect(species, 'oxytoca')~'CDI-similar',
                             str_detect(species, 'aureus')~'CDI-similar',
                             str_detect(species, 'amalonaticus')~'CDI-similar',
                             str_detect(species, 'aeruginosa')~'CDI-similar',
                             str_detect(species, 'cloacae')~'CDI-similar',
                             str_detect(species, 'difficile')~'Cdiff',
                             effect.size > 0.9 ~ 'enriched-relevant',
                             effect.size > 0.1 & effect.size <= 0.9 & p.adj < 0.01 ~ 'enriched',
                             effect.size < -0.1 & p.adj < 0.01 ~ 'depleted',
                             TRUE~'other')) %>% 
  mutate(label=case_when(highlight%in%c('CDI-similar', 'Cdiff', 'enriched-relevant')~
                           str_remove(species, '\\[(ref|meta)_mOTU_v2_[0-9]{4}\\]'),
                         TRUE~'')) %>% 
  mutate(color_positive=p.adj < 0.01 & effect.size > 0.125) %>% 
  mutate(highlight=factor(highlight, 
                          levels = c('Cdiff', 'CDI-similar', 'enriched', 'enriched-relevant',
                                     'depleted', 'other'))) %>% 
  ggplot(aes(x=effect.size, y=-log10(p.adj))) + 
    geom_hline(yintercept = 2, colour='darkgrey', lty=2) + 
    geom_vline(xintercept = c(-0.1, 0.1), colour='darkgrey', lty=2) + 
    geom_point(col='lightgrey') + 
    geom_point(data=. %>% filter(highlight!='other'),
               aes(col=highlight)) +
    geom_text_repel(aes(label=label)) + 
    theme_bw() + 
    theme(panel.grid.minor=element_blank()) + 
    xlab('LME model effect size') + 
    ylab('-log10(adj. P-value)') + 
    scale_colour_manual(
      values=c('#E40046', '#8246AF', '#FFA300', '#FFA300', '#307FE2', 'lightgrey'), 
      labels=c('C. difficile', 'CDI-similar', 'CDI-enriched', 'CDI-enriched-relevant',
               'CDI-depleted','other'), name='')

g2

ggsave(g2, filename = here("figures/CDI/", 'cdiff_lme_volcano_withoutCdiff.pdf'),
       width = 10, height = 8,
       useDingbats=FALSE)
print(g2)
```

# Session info

```{r session_info}
sessionInfo()
```
