---
title: "Species richness and evenness calculations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load tidyverse first
library(tidyverse)
library(vegan)
library(ggpubr)
library(gridExtra)
library(grid)
library("coin")
```

#Load mOTU counts table
```{r }
count_table<-read.csv("../../data/mOTU_counts.tsv", sep = '\t', row.names=1)
```

#filter counts table
```{r}
# remove -1 (unclassified) line
count_table <- count_table[setdiff(rownames(count_table), "-1"), ]

#remove species that never occurr across all samples
count_table_filtered <- count_table[rowSums(count_table) > 0, ]
count_table_filtered_t<-data.frame(t(count_table_filtered))
```

#load metadata table (after timeseries dereplication)
```{r}
meta.data<-read.csv("../../data/metadata_timeseries_dereplicated.csv", sep = ',', stringsAsFactors=F)
rownames(meta.data) <- as.character(meta.data$Samples)
```

#Calculate alpha diversity
```{r }
filename1 <- "alpha_diversity.R"
filename2 <- "rarefaction.R"

#download R scripts sourced if they are not already present
if ((!file.exists(filename1)) | (!file.exists(filename2))) {
  dir.create(path = "external_scripts")
  download.file(url = "https://raw.githubusercontent.com/defleury/Toolbox_16S/master/R/function.alpha_diversity.R", destfile = "external_scripts/alpha_diversity.R")
  download.file(url = "https://raw.githubusercontent.com/defleury/Toolbox_16S/master/R/function.rarefaction.R", destfile = "external_scripts/rarefaction.R")
}

source("external_scripts/alpha_diversity.R")
source("external_scripts/rarefaction.R")

#get as R object the alpha diversity results of already computed. 
filename <- "../../data/alpha_diversity/HillDiv_counts_withMetadata.rds"

if (file.exists(filename)) {
  alpha.frame.m <- readRDS(file = filename)
} else {
    
  #Calculate rarefied alpha diversities
  #alpha div measures are replaced by NA for undersampled samples (see script)
  alpha.frame <- Hill_Diversity.rarefied(
    count_table_filtered, 
    size=100, # min 100 reads per sample
    iterations=100, 
    q.H=c(0, 1, 2) #Hill's numbers
    )
  
  colnames(alpha.frame) = c("Sample", "Richness", "exp_Shannon", "inv_Simpson") 
  #calculate evenness as exp(Shannon)/Richness
  alpha.frame[, "evenness (q1)"] <- alpha.frame$exp_Shannon/ alpha.frame$Richness
  colnames(alpha.frame) = c("Sample", "Richness", "exp_Shannon", "inv_Simpson", "evenness") 
  
  alpha.frame <- alpha.frame[-c(6)]
  rownames(alpha.frame) <- as.character(alpha.frame$Sample)
  
  #Append metadata to alpha diversity results table
  alpha.frame.m <- cbind(meta.data, alpha.frame[rownames(meta.data), ])

  #save as rds object the result of the alpha diversity calculation
  saveRDS(alpha.frame.m, file = filename)
}
```

#remove NA from alpha div values (from undersampled samples)
```{r}
alpha.frame.m <- alpha.frame.m %>% 
  filter(is.na(Richness)==FALSE)
```

#function to count number of samples per each boxplot
```{r}
stat_box_data <- function(y) {
  return(data.frame(y = 0, label = paste(length(y))))
}
```

#define color palette and text size
```{r}
size_text = 10
age_colors <- c("#01579B", "#F9A825")
```


#Combined plots on alpha diversity across age groups and health status
```{r}
library(ggpubr)
library(rstatix)

#plotting function
create_boxplot_grid_age <- function(input.data, x, y, cdiff, status){
    ggplot(data = input.data, mapping = aes(x = !!as.name(x), y = !!as.name(y), fill = !!as.name(cdiff))) +
    geom_boxplot(outlier.size = -1, position = position_dodge(preserve = "single")) +
    geom_jitter(position=position_jitterdodge(jitter.width=0.25), alpha = 0.1, size = 0.05) +
    theme_bw() +
    labs(x="Age group", y = y)+
    theme(plot.title = element_text(family = "Helvetica", face = "bold", size = (10)), 
          #axis.text.x = element_text(angle=45, vjust=1.0, hjust = 1.0), 
          strip.background = element_rect(color="white", fill="white"), 
          axis.title.x = element_text(size= size_text),
          axis.title.y = element_text(size= size_text)) +
    facet_grid(cols = vars(factor(!!as.name(status), levels = c("Healthy", "Diseased"))))+
    scale_fill_manual(values = age_colors) +
    stat_compare_means(aes(group = !!as.name(cdiff)), label = "p.signif") +
    stat_summary(fun.data = stat_box_data, geom = "text", fun = median, position = position_dodge(width = 0.85), size=2)
}

#select human gut stool representative samples, for which age group AND status are known, which studies have at least 2 cdiff+ samples
alpha.frame.m.filtered <- alpha.frame.m %>% 
  filter(Host == "Human" & Environment == "Gut" & Gut_sample_type == "Stool" & final_representative == "Yes") %>% 
  filter(!is.na(Age_group) & !is.na(Status)) %>% 
  #relevel BEFORE grouping
  mutate(Age_group = fct_relevel(Age_group, "Infant", "Child", "Adolescent", "Adult", "Elderly")) %>% 
  group_by(Study) %>% 
  #condition on at least 2 cdiff+ sample has to be AFTER other conditions are met
  dplyr::mutate(atLeastTwoPosCdiff = sum(Cdiff=="Yes")>1) %>% 
  filter(atLeastTwoPosCdiff == TRUE) %>% 
  mutate(age_status_cdiff = paste0(Age_group, "_", Status, "_", Cdiff)) %>% 
  ungroup() %>% 
  group_by(age_status_cdiff) %>% 
  #condition on min samples per boxplot to show it
  dplyr::mutate(n_perBoxplot = dplyr::n()) %>% 
  filter(n_perBoxplot >3) %>% #threhsold of min 3 samples per boxplot
  ungroup()

#non paired t-test 
stat.test <- compare_means(Richness ~ Cdiff, data = alpha.frame.m.filtered,
              group.by = "Age_group", method = "t.test")

#adjusted p values (as stat_compare_means provides only p-value significance)
stat.test <- alpha.frame.m.filtered %>%
  filter(Status=="Healthy" & Age_group=="Infant") %>% 
  #group_by(Status, Cdiff) %>%
  t_test(Richness ~ Cdiff) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance("p.adj")
stat.test

#plot richness
p1 <- create_boxplot_grid_age(alpha.frame.m.filtered, "Age_group", "Richness", "Cdiff",  "Status")

#plot Evenness (for suppl.)
p2 <- create_boxplot_grid_age(alpha.frame.m.filtered, "Age_group", "evenness", "Cdiff", "Status") 

pdf(file="../../figures/alpha_diversity/richness_age.pdf", width=8, height=3)
grid.arrange(p1, nrow = 1)
dev.off()
```

#alpha diversity across age groups and health status - blocked by Study
```{r}
alpha.frame.m.filtered$Study <- as.factor(alpha.frame.m.filtered$Study)
alpha.frame.m.filtered$Cdiff <- as.factor(alpha.frame.m.filtered$Cdiff)

age_group_list <- c("Infant", "Child", "Adolescent", "Adult", "Elderly")
status_list <- c("Healthy", "Diseased")

p.val_vector <- c()
collect.result <- data.frame()

for (age in age_group_list) {
  for  (status in status_list) {

    if((age == "Adolescent"  | age == "Elderly") & status == "Healthy") {next} #not enough data
    print(age)
    print(status)
    
    (subset <- alpha.frame.m.filtered %>%
      filter(Age_group == age & Status == status) %>%
      group_by(Study) %>%
      dplyr::mutate(n= dplyr::n()) %>%
      filter(n > 1) %>% #filter out studies with less than 2 samples
      ungroup())

    #y is richness, x are CDI vs control, blocking by study
    p.val <- pvalue(coin::wilcox_test(Richness ~ Cdiff | Study, data=subset))
    p.val_vector <- c(p.val_vector, p.val)
    print(p.val)
    print("****")

    collect.result <- rbind(collect.result, data.frame(age_group = age, status = status, pvalue = p.val))
  }
}
```

# Combined plots on alpha diversity across delivery mode and health status 
```{r}
#plotting function
create_boxplot_grid_birthMode_prematurity <- function(input.data, x, y, cdiff, status_term){
ggplot(data = input.data, mapping = aes(x = !!as.name(x), y = !!as.name(y), fill = !!as.name(cdiff))) +
    geom_boxplot(outlier.size = -1) +
    theme_bw() +
    labs(x="Delivery mode", y = y)+
    theme(strip.background = element_rect(color="white", fill="white"),
          axis.title.x = element_text(size= size_text),
          axis.title.y = element_text(size= size_text)) +
    facet_grid(cols = vars(factor(!!as.name(status_term), levels = c("Premature_Healthy", "Premature_Diseased", "Full-Term_Healthy", "Full-Term_Diseased")))) +
    scale_fill_manual(values = age_colors) +
    stat_compare_means(aes(group = !!as.name(cdiff)), label = "p.signif")+
    stat_summary(fun.data = stat_box_data, geom = "text", fun = median, position = position_dodge(width = 0.85), size=2)
}

#select human gut stool representative samples from PREMATURE babies, for which age group AND status are known, which studies have at least 2 cdiff+ samples
alpha.frame.m.filtered.birth.mode.premature <- alpha.frame.m %>% 
  filter(Host == "Human" & Environment == "Gut" & Gut_sample_type == "Stool" & final_representative == "Yes" & Age_group == "Infant") %>% 
  filter(!is.na(Delivery_mode) & !is.na(Status) & !is.na(Premature)) %>% 
  mutate (Premature_extended = ifelse(Premature == "Yes", "Premature", "Full-Term"),
          #relevel BEFORE grouping
          Delivery_mode = fct_relevel(Delivery_mode, "Vaginal", "Cesarean")) %>% 
  group_by(Study) %>% 
  #condition on at least 2 cdiff+ sample has to be AFTER other conditions are met
  dplyr::mutate(atLeastTwoPosCdiff = sum(Cdiff=="Yes")>1) %>% 
  filter(atLeastTwoPosCdiff == TRUE) %>% 
  mutate(status_term = paste0(Premature_extended, "_", Status))

stat.test <- compare_means(
 Richness ~ Delivery_mode, data = alpha.frame.m.filtered.birth.mode.premature,
 method = "t.test"
)

#plot richness
p3 <- create_boxplot_grid_birthMode_prematurity(alpha.frame.m.filtered.birth.mode.premature, 
                                                "Delivery_mode", "Richness", "Cdiff", "status_term")

#plot Evenness
p4 <- create_boxplot_grid_birthMode_prematurity(alpha.frame.m.filtered.birth.mode.premature, 
                                                "Delivery_mode", "evenness", "Cdiff","status_term")

#plot exp_Shannon
p5 <- create_boxplot_grid_birthMode_prematurity(alpha.frame.m.filtered.birth.mode.premature, 
                                                "Delivery_mode", "exp_Shannon", "Cdiff", "status_term")

a <- alpha.frame.m %>% 
  filter(Host == "Human" & Environment == "Gut" & Gut_sample_type == "Stool" & final_representative == "Yes" & Age_group == "Infant") %>% 
  filter(Status=="Diseased" & Premature=="No")
```

#alpha diversity across prematurity and delivery mode - blocked by Study
```{r}
alpha.frame.m.filtered.birth.mode.premature$Study <- as.factor(alpha.frame.m.filtered.birth.mode.premature$Study)
alpha.frame.m.filtered.birth.mode.premature$Cdiff <- as.factor(alpha.frame.m.filtered.birth.mode.premature$Cdiff)

delivery_mode_list <- c("Vaginal", "Cesarean")
status_term_list <- c("Premature_Healthy", "Premature_Diseased", "Full-Term_Healthy")

collect.result.delivery.term <- data.frame()

for (delivery_mode in delivery_mode_list) {
  for  (status_term in status_term_list) {

    subset <- alpha.frame.m.filtered.birth.mode.premature %>%
      filter(Delivery_mode == delivery_mode & status_term == status_term) %>%
      group_by(Study) %>%
      dplyr::mutate(n= dplyr::n()) %>%
      filter(n > 1) %>% #filter out studies with less than 2 samples
      ungroup()

    #y is richness, x are CDI vs control, blocking by study
    p.val <- pvalue(coin::wilcox_test(Richness ~ Cdiff | Study, data=subset))
    p.val_vector <- c(p.val_vector, p.val)
    print(p.val)
    print("****")

    collect.result.delivery.term <- rbind(collect.result.delivery.term, data.frame(Delivery_mode = delivery_mode, status_term = status_term, pvalue = p.val))
  }
}
```

#alpha diversity by age groups and by country
```{r}
create_boxplot_grid_geography_age <- function(input.data, x, y, cdiff, Age_group){
ggplot(data = input.data, mapping = aes_string(x = x, y = y, fill = cdiff)) +
    geom_boxplot(outlier.size = -1) +
    geom_jitter(position=position_jitterdodge(jitter.width=0.25), alpha = 0.2, size = 0.05) +
    theme_bw() +
    labs(x="Continent", y = y)+
    theme(strip.background = element_rect(color="white", fill="white"), 
          axis.text.x = element_text(angle=0, vjust=1.0, hjust=1.0), 
          axis.title.x = element_text(size= size_text),
          axis.title.y = element_text(size= size_text)) +
    facet_grid(cols = vars(factor(!!as.name(Age_group), levels = c("Infant")))) +
    scale_fill_manual(values = age_colors) +
    stat_compare_means(aes_string(group = cdiff), label = "p.signif")+
    stat_summary(fun.data = stat_box_data, geom = "text", fun = median, position = position_dodge(width = 0.85), size=2)
}

alpha.frame.m.filtered.continentAgeGroups <- alpha.frame.m %>% 
  filter(Host == "Human" & Environment == "Gut" & Gut_sample_type == "Stool" & final_representative == "Yes") %>% 
  filter(!is.na(Continent) & Status == "Healthy" & (Age_group == "Infant")) %>% 
  group_by(Continent, Age_group, Cdiff) %>% 
  #condition on at least 5 cdiff+ sample has to be AFTER other conditions are met
  dplyr::mutate(nsamples = dplyr::n()) %>% 
  filter(nsamples>4 & Continent != "Africa") 

stat.test <- compare_means(Richness ~ Continent, data = alpha.frame.m.filtered.continentAgeGroups, method = "t.test")
p6 <- create_boxplot_grid_geography_age(alpha.frame.m.filtered.continentAgeGroups, "Continent", "Richness", "Cdiff", "Age_group")
```

#alpha diversity across continent - blocked by Study
```{r}
alpha.frame.m.filtered.continentAgeGroups$Study <- as.factor(alpha.frame.m.filtered.continentAgeGroups$Study)
alpha.frame.m.filtered.continentAgeGroups$Cdiff <- as.factor(alpha.frame.m.filtered.continentAgeGroups$Cdiff)

age_group_list <- c("Infant")
continent_list <- c("Asia", "Europe", "North America", "Oceania")

collect.result.continent <- data.frame()

for (age_group in age_group_list) {
  for  (continent in continent_list) {

    if(age_group == "Adult" & continent == "Oceania") {next} #not enough data
    print(age_group)
    print(continent)
    
    subset <- alpha.frame.m.filtered.continentAgeGroups %>%
      filter(Age_group == age_group & Continent == continent) %>%
      group_by(Study) %>%
      dplyr::mutate(n= dplyr::n()) %>%
      filter(n > 1) %>% #filter out studies with less than 2 samples
      ungroup()

    #y is richness, x are CDI vs control, blocking by study
    p.val <- pvalue(coin::wilcox_test(Richness ~ Cdiff | Study, data=subset))
    p.val_vector <- c(p.val_vector, p.val)
    print(p.val)
    print("****")

    collect.result.continent <- rbind(collect.result.continent, data.frame(Age_group = age_group, Continent = continent, pvalue = p.val))
  }
}
```

#plot species richness by delivery mode, prematurity and geography + species eveness
```{r}
pdf(file="../../figures/alpha_diversity/richness_evenness_suppl.pdf", width=10, height=15)
grid.arrange(p2, p3, p4, p6, ncol = 1)
dev.off()
```

